---
title: Proteomics QC Report
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
params:
    data: merged.mzTab
---

```{r include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
library(MSnbase)
library(ggplot2)
library(tidyr)
library(knitr)
library(pheatmap)
library(grid)
## ggplot theme:
theme_set(theme_bw() + theme(plot.title=element_text(hjust=0.5, face="bold"), plot.subtitle=element_text(hjust=0.5)))

data <- MzTab(params$data)
meta <- data@Metadata
psms <- data@PSMs
pep <- data@Peptides
prot <- data@Proteins
inputs <- unlist(meta[grep("ms_run\\[\\d+\\]-location", names(meta))])
```

`r Sys.time()`

## General information

Input file: `r normalizePath(params$data)`

**`r length(inputs)`** samples processed:

```{r results="asis"}
samples <- sub("file://(.*)\\.[^\\.]+", "\\1", inputs)
cat(paste("#.", samples), sep="\n")
```

## Identification level

Max. q-value (PSM level): `r round(max(psms$"opt_global_q-value", na.rm=TRUE), 3)`

```{r fig.show="hold", fig.width=4, fig.height=4, out.width="50%"}
## count MS2 spectra:
n.ms2 <- as.integer(unlist(meta[grep("ms_run\\[\\d+\\]-num_ms2", names(meta))]))
## count PSMs:
run.ids <- as.integer(sub("^ms_run\\[(\\d+)\\]:.*", "\\1", psms$spectra_ref))
n.psms <- table(factor(run.ids, levels=seq_along(inputs)))

id.rates <- as.numeric(n.psms / n.ms2)
plot.data <- data.frame(sample=factor(names(n.psms), levels=seq_along(inputs)),
                        "MS2 spectra"=n.ms2, "PSMs"=as.integer(n.psms),
                        "id. rate (%)"=id.rates * 100, check.names=FALSE)

ggplot(plot.data) +
  geom_col(aes(sample, `MS2 spectra`)) +
  labs(title="Number of MS2 spectra acquired")

ggplot(plot.data) +
  geom_col(aes(sample, `id. rate (%)`)) +
  labs(title="Fraction of MS2 spectra identified", subtitle="(subject to q-value threshold)")
```

```{r}
counts <- plot.data[, c("sample", "PSMs")]
names(counts)[2] <- "count"
counts$quantity <- "PSMs"
## count peptides:
cols <- paste0("search_engine_score[1]_ms_run[", seq_along(inputs), "]")
n.peps <- apply(pep[, cols], 2, function(col) sum(!is.na(col)))
counts <- rbind(counts, data.frame(count=n.peps, sample=seq_along(n.peps),
                                   quantity="charged peptides\n(incl. mods)"))
## unique peptides:
n.unique <- sapply(cols, function(col) {return
    part <- pep[!is.na(pep[[col]]), c("sequence", "modifications")]
    c(nrow(unique(part)), length(unique(part$sequence)))
})
n.unique <- data.frame(count=as.integer(n.unique), sample=rep(seq_along(inputs), each=2),
                       quantity=c("unique peptides\n(incl. mods)", "unique peptides\n(w/o mods)"))
counts <- rbind(counts, n.unique)
## count proteins:
cols <- paste0("search_engine_score[1]_ms_run[", seq_along(inputs), "]")
## don't count proteins inside ambiguity groups ("ambiguity_members" is NA):
n.prots <- apply(prot[!is.na(prot$ambiguity_members), cols], 2, function(col) sum(!is.na(col)))
counts <- rbind(counts, data.frame(count=n.prots, sample=seq_along(n.peps),
                                   quantity="protein groups"))
## ensure proper ordering:
counts$sample <- factor(counts$sample, levels=seq_along(inputs))
counts$quantity <- factor(counts$quantity, levels=c("PSMs", "charged peptides\n(incl. mods)",
                                                    "unique peptides\n(incl. mods)", "unique peptides\n(w/o mods)",
                                                    "protein groups"))

ggplot(counts) +
    geom_col(aes(sample, count, fill=quantity), position="dodge") +
    labs(title="PSM/peptide/protein counts")
```

```{r results="asis"}
table <- pivot_wider(counts, names_from="quantity", values_from="count")
table <- cbind(plot.data[, -3], table[, -1])
names(table) <- gsub("\n", "<br/>", names(table)) # fix line breaks for HTML output
kable(table, digits=1) # caption="Table of identification-related metrics"
```

### Chromatography

```{r message=FALSE, warning=FALSE}
## retention times:
min.samples <- 3 # min. number of samples a peptide must be identified in to be used
outlier.pct <- 1 # percentage of outliers to be removed
rt.data <- psms[, c("sequence", "modifications", "retention_time")]
rt.data$sample <- factor(run.ids, levels=seq_along(inputs))
rt.data <- pivot_wider(rt.data, names_from=sample, values_from=retention_time, values_fn=median)
n.samples <- apply(rt.data[, -(1:2)], 1, function(row) sum(!is.na(row)))
rt.part <- rt.data[n.samples >= min.samples, ]
rt.part$median_rt <- apply(rt.part[, -(1:2)], 1, median, na.rm=TRUE)
rt.part <- pivot_longer(rt.part, as.character(seq_along(inputs)), names_to="sample", values_to="rt",
                        values_drop_na=TRUE)
rt.part$rt_diff <- rt.part$rt - rt.part$median_rt
rt.part$sample <- factor(rt.part$sample, levels=seq_along(inputs))
outlier.frac <- outlier.pct / 200
ggplot(rt.part, aes(median_rt, rt_diff, color=sample)) +
    geom_point(shape=".") +
    geom_smooth(size=0.5) +
    ylim(quantile(rt.part$rt_diff, c(outlier.frac, 1 - outlier.frac))) +
    labs(x="median RT (s)", y="RT deviation (s)", title="Retention times of identified peptides")
```

The median retention time (RT) of a peptide in each sample is compared to its median RT across all samples.
Only peptides identified in at least `r min.samples` samples are considered.
For scaling purposes, `r outlier.pct`% of outliers in terms of RT deviation were excluded from the plot.


## Quantification level

### Peptides

```{r}
plot.abundance.distributions <- function(data.part, col.pattern, title, subtitle=NULL) {
  ab.cols <- grep(col.pattern, names(data.part), value=TRUE)
  ab.data <- data.part[, ab.cols]
  names(ab.data) <- sub(col.pattern, "\\1", ab.cols)
  plot.data <- pivot_longer(ab.data, 1:ncol(ab.data), names_to="sample", values_to="abundance",
                            values_drop_na=TRUE)
  plot.data$sample <- factor(plot.data$sample, levels=seq_along(inputs))
  plot <- ggplot(plot.data) +
    geom_density(aes(log10(abundance), color=sample)) +
    labs(title=title, subtitle=subtitle)
  print(plot)
  invisible(ab.data)
}

plot.abundance.distributions(pep, "^peptide_abundance_study_variable\\[(\\d+)\\]$",
                             "Distribution of peptide abundances", "(incl. charges and modifications)")
```

### Proteins

```{r}
ind <- prot$opt_global_result_type != "protein_details"
prot.ab <- plot.abundance.distributions(prot[ind, ], "^protein_abundance_study_variable\\[(\\d+)\\]$",
                                        "Distribution of protein (group) abundances")

## heatmap:
prot.ab <- as.matrix(prot.ab)
rownames(prot.ab) <- paste0(prot$accession[ind],
                            ifelse(prot$opt_global_result_type[ind] == "indistinguishable_protein_group", "*", ""))
n.samples <- apply(prot.ab, 1, function(row) sum(!is.na(row)))
min.samples <- ceiling(ncol(prot.ab) / 2) # require quantification in at least half the samples
ind <- n.samples >= min.samples
plot.data <- log10(prot.ab[ind, ])
ord <- order(rowSums(plot.data, na.rm=TRUE), decreasing=TRUE)
heatmap_height <- sum(ind) / 10
```

### Heatmap of protein abundances (log-scale)

```{css}
.with-scrollbar {
    overflow-y: scroll;
    max-height: 400px;
}
```

<div class="with-scrollbar">
```{r fig.height=heatmap_height}
## using chunk option 'class.output="with-scrollbar"' (instead of '<div>') doesn't work for some reason

## just calling "pheatmap" doesn't produce output in the HTML file:
p <- pheatmap(plot.data[ord, ], cluster_rows=FALSE, cluster_cols=FALSE, angle_col=0, fontsize_row=6)
grid.draw(p$gtable)
```
</div>

<br/>
Abundances are shown for **`r nrow(plot.data)`** proteins (or ambiguity groups) that were quantified in at least half of all samples (i.e. min. `r min.samples` samples).
